---
title: "Initial Analysis: Thesis"
author: "Vir Chachra"
date: "`r Sys.Date()`"
output: html_document
---


```{r datasets}

library(tidyverse)
library(tidycensus)
library(data.table)
library(sf)
library(kableExtra)
library(ggspatial)
library(ggmap)
library(gt)
library(stargazer)
library(rempsyc)
library(vtable)
library(findSVI)
library(readr)
library(data.table)
library(rgeoda)
library(writexl)
library(readxl)
colors <- c("#0073C2FF", "#EFC000FF", "#868686FF", "#CD534CFF", "#7AA6DCFF", "#388E3CFF", "#D32F2FFF")
state_list <- c("MA", "RI", "IL", "IN", "PA", "NJ", "CA", "WA", "NY", "CT", "GA")

# Census race data pull 

MBTA_tracts_21 <- st_read("2021_MBTA_DOT_Index.geojson") %>%  select(GEOID, mode_group, cr_tracts, bus_tracts, rt_tracts)  %>% mutate(operator = "MBTA")

SEPTA_tracts_21 <- st_read("2021_SEPTA_DOT_Index.geojson") %>%  select(GEOID, mode_group, cr_tracts, bus_tracts, rt_tracts) %>% mutate(operator = "SEPTA")

Chicago_tracts_21 <- st_read("2021_CTA_DOT_Index.geojson") %>%  select(GEOID, mode_group, cr_tracts, bus_tracts, rt_tracts)  %>% mutate(operator = "Chicago")


LA_tracts_21 <- st_read("2021_LA_DOT_Index.geojson") %>%  select(GEOID, mode_group, cr_tracts, bus_tracts, "rt_tracts" = hr_tracts) %>% mutate(operator = "LA")

Seattle_tracts_19 <- st_read("2019_Seattle_DOT_Index.geojson") %>%  select(GEOID, mode_group, cr_tracts, bus_tracts, rt_tracts)  %>% mutate(operator = "Seattle")
Seattle_tracts_21 <- st_read("2021_Seattle_DOT_Index.geojson") %>%  select(GEOID, mode_group, cr_tracts, bus_tracts, rt_tracts)  %>%  mutate(operator = "Seattle")

MTA_tracts_21 <- st_read("2021_MTA_DOT_Index.geojson") %>%  select(GEOID, mode_group, cr_tracts, bus_tracts, rt_tracts)  %>%  mutate(operator = "MTA")

MARTA_tracts_21 <- st_read("2021_MARTA_DOT_Index.geojson") %>%  select(GEOID, mode_group, bus_tracts, rt_tracts)  %>%  mutate(operator = "MARTA") %>% mutate(cr_tracts = 0)



colors <- c("#0073C2FF", "#EFC000FF", "#868686FF", "#CD534CFF", "#7AA6DCFF", "#388E3CFF", "#D32F2FFF", "#FF5722FF", "#8E44ADFF")


#Add transit miles, transit spending, and transportation cost burdened population from 2019 and 2021 TIAT datasets, do it once and then
#comment out the code

#Adjusting for inflation to June 2021 dollars

inflation_2019_2021 = 1.06

tiat_tracts <- st_read("DOT TIAT Source Data//TIAT_Cost_Burden_Tracts_2021_v01_2024_11_24//tracts_2021//cb_2020_us_tract_500k//cb_2020_us_tract_500k.shp")


tiat_2019 <- fread("profile_final_tract_2019.csv") %>%
  select(
    geoid,
    state_name,
    hh_profile,
    transit_spending,
    median_income,
    mean_income,
    population,
    employment_density,
    transportation_cost,
    transportation_cost_burden,
    transit_miles
  ) %>%
  mutate(
    inflation_adjusted_transport_cost = inflation_2019_2021 * transportation_cost,
    inflation_adjusted_transit_cost = inflation_2019_2021 * transit_spending,
    inflation_adjusted_median_income = inflation_2019_2021 * median_income, 
    inflation_adjusted_mean_income = inflation_2019_2021 * mean_income, 
    inflation_adjusted_transportation_cost_burden = (inflation_adjusted_transport_cost / inflation_adjusted_median_income),
    inflation_adjusted_transit_cost_burden = (inflation_adjusted_transit_cost / inflation_adjusted_median_income),
    
    
    log_transportation_cost_burden = log(inflation_adjusted_transportation_cost_burden),
    log_transit_cost_burden = log(inflation_adjusted_transit_cost_burden),
    log_population = log(population),
    log_transit_miles = log(transit_miles),
    log_emp_density = log(employment_density),
    log_transport_cost = log(inflation_adjusted_transport_cost), 
    log_transit_spending = log(inflation_adjusted_transit_cost)
  )

# have to standardize and convert geoids
checks_2019 <- data.frame("original_geoid" = unique(tiat_2019$geoid)) %>%
  mutate(char_geoid = as.character(original_geoid),
         nchars = nchar(char_geoid))

checks_2019 <- checks_2019 %>%
  mutate(char_geoid = ifelse(nchars == 10, paste0("0", char_geoid), char_geoid)) %>%
  mutate(nchars = nchar(char_geoid))


tiat_2019 <- tiat_2019 %>%
  left_join(checks_2019 %>% select(original_geoid, char_geoid),
            by = c("geoid" = "original_geoid")) %>% mutate(geoid = char_geoid) %>% select(-char_geoid)




# Read and process 2021 dataset
tiat_2021 <- fread("profile_tract_2021.csv") %>%
 select(
    geoid,
    state_name,
    hh_profile,
    transit_spending,
    median_income,
    mean_income,
    population,
    employment_density,
    transportation_cost,
    transportation_cost_burden,
    transit_miles
  ) %>%
  mutate(
    inflation_adjusted_transport_cost = transportation_cost,
    inflation_adjusted_transit_cost =  transit_spending,
    inflation_adjusted_median_income =  median_income, 
    inflation_adjusted_mean_income = mean_income, 
    inflation_adjusted_transportation_cost_burden = (inflation_adjusted_transport_cost / inflation_adjusted_median_income),
    inflation_adjusted_transit_cost_burden = (inflation_adjusted_transit_cost / inflation_adjusted_median_income),
    
    
    log_transportation_cost_burden = log(inflation_adjusted_transportation_cost_burden),
    log_transit_cost_burden = log(inflation_adjusted_transit_cost_burden),
    log_population = log(population),
    log_transit_miles = log(transit_miles),
    log_emp_density = log(employment_density),
    log_transport_cost = log(inflation_adjusted_transport_cost), 
    log_transit_spending = log(inflation_adjusted_transit_cost)
  )


checks_2021 <- data.frame("original_geoid" = unique(tiat_2021$geoid)) %>%
  mutate(char_geoid = as.character(original_geoid),
         nchars = nchar(char_geoid))

checks_2021 <- checks_2021 %>%
  mutate(char_geoid = ifelse(nchars == 10, paste0("0", char_geoid), char_geoid)) %>%
  mutate(nchars = nchar(char_geoid))


tiat_2021 <- tiat_2021 %>%
  left_join(checks_2019 %>% select(original_geoid, char_geoid),
            by = c("geoid" = "original_geoid")) %>% mutate(geoid = char_geoid) %>% select(-char_geoid)



# tiat_tracts_w_data <- tiat_tracts %>% left_join(tiat_2019, by = c("GEOID" = "geoid")) %>% left_join(tiat_2021, by = c("GEOID"="geoid"))

#CDC SVI - Using this for consistency instead (change 2.16.2025)




# #svi needs to be crosswalked to 2021 geospatial data
# svi_2019 <- find_svi(
#   year = rep(2019, length(state_list)),
#   state = state_list,
#   geography = "tract",
# )
# 
census_crosswalk <- read.csv("census_crosswalk.csv")

# Crosswalk to standardize geographies for join
# svi_2019_crosswalked <- svi_2019 %>%
#   mutate(GEOID = as.numeric(GEOID)) %>%
#   left_join(census_crosswalk %>% select("GEOID" = GEOID_TRACT_10, "GEOID_NEW" = GEOID_TRACT_20), by = "GEOID") %>%
#   group_by(GEOID_NEW) %>%
#   summarise(RPL_themes = mean(RPL_themes),
#             RPL_theme1 = mean(RPL_theme1),
#             RPL_theme2 = mean(RPL_theme2),
#             RPL_theme3 = mean(RPL_theme3),
#             RPL_theme4 = mean(RPL_theme4)) %>%
#   mutate(char_geoid = as.character(GEOID_NEW),
#          nchars = nchar(char_geoid))
# 
# svi_2019_crosswalked <- svi_2019_crosswalked %>%
#    mutate(char_geoid = ifelse(nchars == 10, paste0("0", char_geoid), char_geoid)) %>%
#   mutate(nchars = nchar(char_geoid),
#          GEOID = char_geoid) %>%
#   select(-c(GEOID_NEW, char_geoid, nchars))
# 
# 
# 
# 
# 
# svi_2021 <- find_svi(
#   year =  rep(2021, length(state_list)),
#   state = state_list,
#   geography = "tract"
# )

# writexl::write_xlsx(svi_2019_crosswalked, "svi_2019.xlsx")
# writexl::write_xlsx(svi_2021, "svi_2021.xlsx")

svi_2019 <- read_xlsx("svi_2019.xlsx")
svi_2021 <- read_xlsx("svi_2021.xlsx")

# svi_2019 <- read.csv("svi_2019.csv")
# svi_2021 <- read.csv("svi_2021.csv")


#Seattle Extension is only difference 
transit_tracts_2019 <- rbind(MBTA_tracts_21,
                            LA_tracts_21,
                            Chicago_tracts_21,
                            SEPTA_tracts_21,
                            Seattle_tracts_19,
                            MTA_tracts_21,
                            MARTA_tracts_21)


transit_tracts_2021 <- rbind(MBTA_tracts_21,
                            LA_tracts_21,
                            Chicago_tracts_21,
                            SEPTA_tracts_21,
                            Seattle_tracts_21,
                            MTA_tracts_21,
                            MARTA_tracts_21)

combined_data_2019 <- transit_tracts_2021 %>%
  left_join(tiat_2019 %>% filter(hh_profile == 1), by = c("GEOID" = "geoid")) %>%
  left_join(svi_2019 %>% rename("si_2019" = RPL_themes), by = "GEOID") %>%
  mutate(across(where(is.numeric), ~ ifelse(is.infinite(.), log(pmax(
    ., 1e-6
  )), .))) %>% st_as_sf()

combined_data_2021 <- transit_tracts_2021 %>%
  left_join(tiat_2021 %>% filter(hh_profile == "1: Average Household"), by = c("GEOID" = "geoid")) %>%
  left_join(svi_2021 %>% rename("si_2021" = RPL_themes) %>% select(-year, -state) , by = "GEOID") %>%
  mutate(across(where(is.numeric), ~ ifelse(is.infinite(.), log(pmax(
    ., 1e-6
  )), .))) %>% st_as_sf()


 

full_data <- rbind(
  combined_data_2019 %>% mutate(Year = "2019") %>% rename("si_index" = si_2019),
  combined_data_2021 %>% mutate(Year = "2021") %>% rename("si_index" = si_2021)
) %>% 
  mutate(mode_group = case_when(
    mode_group == "Only RT Access" ~ "Bus and Rapid Transit Access",
    mode_group == "Only Rapid Transit Access" ~ "Bus and Rapid Transit Access",
    mode_group == "CR and Rapid Transit Access" ~ "All Modes",
    TRUE ~ mode_group  # Keeps the original value if no conditions are met
  ),
   Year_2021 = ifelse(Year == 2021, 1,0),
   si_index_mod = ifelse(is.na(si_index), 1, ifelse(si_index == 0, .1, si_index)),
  log_si_index_mod = log(si_index_mod),
  si_index_high = ifelse(si_index > mean(si_index, na.rm = T), 1, 0))



write_sf(full_data, "thesis_tract_data.geojson")



```

#Summary Mode Group Maps 
```{r}
library(tidyverse)
library(tmap)
library(ggplot2)
library(sf)
library(showtext)  # For better font handling
library(ggtext)    # Allows for HTML-rich text formatting
library(patchwork) 



full_data <- read_sf("thesis_tract_data.geojson")





full_data$cr_tracts_only <- ifelse(full_data$mode_group == "Only CR Access", 1, 0)

full_data$log_si_index <- full_data$log_si_index_mod


#TRANSPORT COST BURDEN CORRECTION
full_data_mod <- full_data %>%
  mutate(
    inflation_adjusted_transportation_cost_burden = inflation_adjusted_transport_cost / inflation_adjusted_median_income,
    log_transportation_cost_burden = log(inflation_adjusted_transportation_cost_burden)
  ) %>%
  filter(!is.na(si_index) & !is.na(log_transportation_cost_burden))


#st_write(full_data_mod, "thesis_tract_data.geojson",delete_dsn = T)

full_data <- full_data_mod

full_data$urban_areas <- case_when(
  full_data$operator == "Chicago" ~ "Chicago (CTA, Metra, SSL)",
  full_data$operator == "LA" ~ "Los Angeles (LA Metro)",
  full_data$operator == "MBTA" ~ "Boston (MBTA)",
  full_data$operator == "SEPTA" ~ "Philadelphia (SEPTA)",
  full_data$operator == "Seattle" ~ "Seattle (Sound Transit, King County Metro, Seattle Sounder)",
  full_data$operator == "MTA" ~ "New York City (MTA, Metro-North, LIRR)",
  full_data$operator == "MARTA" ~ "Atlanta (MARTA)"
)


# Load custom fonts (optional)
# Set up font
showtext_auto()
font_add_google("Lato", "lato")

# Filter for 2021
data_2021 <- full_data %>% filter(Year == 2021)
data_2019 <- full_data %>% filter(Year == 2019)
# Set mode
tmap_mode("plot")


 # tm_basemap("OpenStreetMap")  + tm_shape(data_2021) + tm_polygons(
 #    col = "mode_group",
 #    palette = "cividis",
 #    title = "Mode Group"
 #  )

# Create a single faceted map with basemap
showtext_auto()
font_add_google("Lato", "lato")

n_groups <- length(unique(data_2021$mode_group))


# Get consistent palette from cols4all
mode_colors <- cols4all::c4a("cividis", n = n_groups)

  # Use 'plot' for static maps

# Build the map - operators
map <- tm_basemap("CartoDB.PositronNoLabels") +
  tm_shape(data_2021) +
  tm_fill(
    col = "mode_group",
    palette = mode_colors,
    border.col = NULL,
    legend.is.portrait = FALSE,
    title = "Mode Groups",
    na.color = NA
  ) +
  tm_facets(
    by = "urban_areas",
    ncol = 2,
    drop.units = TRUE,
    free.coords = TRUE
  ) +
  tm_layout(
    frame = FALSE,
    bg.color = "white",
    fontfamily = "Bahnschrift",
    legend.outside = TRUE,
    legend.outside.position = "bottom",
    legend.align = "center",
    legend.text.color = "black",
    legend.text.size = 1,
    legend.title.size = 1.3,
    legend.title.fontface = "bold",
    legend.bg.color = NA,
    legend.bg.alpha = 0,
    panel.label.bg.color = "white",
    panel.label.frame = FALSE,
    panel.label.color = "black",
    panel.label.size = 1.4,
    panel.label.height = 1.2,
    panel.label.margin = grid::unit(2, "lines"),
    asp = 0,
#    outer.margins = c(0.1, 0.1, 0.15, 0.1),
  #  inner.margins = 0
  )

tmap_save(
  map,
  filename = "report_faceted_transit_map.pdf",
  width = 12,
  height = 14,
  dpi = 800
)


# Build the map - transit cost burden
map_burden <- tm_basemap("CartoDB.PositronNoLabels") +
  tm_shape(data_2019) +
  tm_fill(
    col = "log_transit_cost_burden",
    palette = "inferno",  # high-contrast, dark-to-light
    border.col = NULL,
    style = "cont",
    legend.is.portrait = FALSE,
    title = "Log 2019 Transit Cost Burden (% of Median Household Income)",
    na.color = NA
  ) +
  tm_facets(
    by = "urban_areas",
    ncol = 2,
    drop.units = TRUE,
    free.coords = TRUE
  ) +
  tm_layout(
    frame = FALSE,
    bg.color = "white",
    fontfamily = "Bahnschrift",
    legend.outside = TRUE,
    legend.outside.position = "bottom",
    legend.align = "center",
    legend.text.color = "black",
    legend.text.size = 1,
    legend.title.size = 1.3,
    legend.title.fontface = "bold",
     legend.stack = "horizontal",
    legend.bg.color = NA,
    legend.bg.alpha = 0,
    panel.label.bg.color = "white",
    panel.label.frame = FALSE,
    panel.label.color = "black",
    panel.label.size = 1.4,
    panel.label.height = 1.2,
    panel.label.margin = grid::unit(2, "lines"),
    asp = 0
  )

tmap_save(
  map_burden,
  filename = "report_faceted_transit_cb.pdf",
  width = 12,
  height = 14,
  dpi = 800
)

library(flextable)

# Create a summary table
table_data <- full_data %>%
  st_drop_geometry() %>%
  group_by(Year, operator, mode_group) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = "operator", values_from = "count")


# Map for eaach operator

library(tmap)
library(purrr)
library(dplyr)

# Ensure tmap is in plot mode
tmap_mode("plot")

# Create a vector of unique operators
unique_operators <- unique(data_2021$operator)

# Loop through each operator and save a map
walk("MBTA", function(op) {
  
  # Filter data for this operator
  op_data <- filter(data_2021, operator == op)
  
  # Create the map
  op_map <- tm_basemap("CartoDB.PositronNoLabels") +
    tm_shape(op_data) +
    tm_fill(
      col = "mode_group",
      palette = mode_colors,
      border.col = NULL,
      legend.is.portrait = FALSE,
      title = "Mode Groups"
    ) +
    tm_layout(
      frame = FALSE,
      bg.color = "#262626",
      
      # Legend appearance
      legend.outside = TRUE,
      legend.outside.position = "bottom",
      legend.align = "center",
      legend.text.color = "white",
      legend.text.size = 0.8,
      legend.title.size = 1,
      legend.title.fontface = "bold",
      legend.bg.color = NA,
      legend.bg.alpha = 0,

      # Title settings (optional)
      main.title = op,
      main.title.size = 1.5,
      main.title.color = "white",
      main.title.fontface = "bold",

      # Layout
      asp = 0,
      outer.margins = 0,
      inner.margins = 0
    )

  # Save the map as a high-resolution PDF
  tmap_save(
    op_map,
    filename = paste0("map_", gsub("[^A-Za-z0-9]", "_", op), ".pdf"),
    width = 8.5,  # can tweak for your publication
    height = 11,
    dpi = 600
  )
})


```
#SVI Index Sample Map


```{r}
library(tmap)

MBTA_2021 <- full_data %>% filter(operator == "MBTA", Year == 2021) %>% filter(!is.na(si_index))

MBTA_CR <- read_sf("C:\\Users\\virch\\MIT Dropbox\\Vir Chachra\\Vir Chachra’s files\\Thesis\\Thesis_Exploration\\Transit Systems Shapefile\\MBTA\\MBTA_Commuter_Rail_Lines\\MBTA_Commuter_Rail_Lines.shp")



earth_palette <- c(
  "#8c6d31", # golden brown
  "#bd925a", # tan
  "#a1b56c", # sage green
  "#7b9e87", # dusty teal
  "#5d7680", # slate blue
  "#3b3b3b"  # deep charcoal
)

# Map with MBTA commuter rail overlay
si_index_map <- tm_shape(MBTA_2021) +
  tm_polygons(
    "si_index",
    palette = earth_palette,
    border.col = NULL
  ) +
  tm_shape(MBTA_CR) +
  tm_lines(
    col = "white",
    lwd = 2
  ) +
  tm_layout(
    title.position = c("left", "top"),
    title.size = 2.5,
    title.color = "black",
    fontfamily = "Bahnschrift",
    frame = FALSE,
    bg.color = NULL,
    legend.show = FALSE,
    inner.margins = c(0.02, 0.02, 0.02, 0.02),
    asp = 0
  )

tmap_save(
  si_index_map,
  filename = "MBTA_2021_SVI_Index_Transparent.png",
  width = 8,
  height = 6,
  dpi = 800,
  bg = NA,  # ensures transparency
  asp = 0
)



```


#Transit and Transport Cost Burden Sample Map - Philly

```{r}
SEPTA_2021 <- full_data %>% filter(operator == "SEPTA", Year == 2021) %>% filter(!is.na(transportation_cost_burden))

SEPTA_2019 <- full_data %>% filter(operator == "SEPTA", Year == 2019) %>% filter(!is.na(transportation_cost_burden))


transport_burden_map_SEPTA <- tm_shape(SEPTA_2021) +
    tm_polygons(
      "transportation_cost_burden",
      palette = "viridis",
      title = "Total Transport Cost Burden 2021 - SEPTA",
      border.col = NULL
    ) + tm_layout(
      title.position = c("left", "top"),
      legend.position = c("right", "bottom"),
      # Dynamic legend placement
      frame = FALSE,
      # No external frame
      bg.color = "white",
      legend.text.size = 1,
      # 12-point font equivalent, adjust for visibility
      legend.title.size = 1.2,
      # Title size\
      legend.outside = TRUE,
      title.color = "black"
    )
 
  transport_burden_map_SEPTA <- transport_burden_map_SEPTA + tm_layout(
    fontfamily = "Bahnschrift",
    legend.title.fontface = "bold")


transit_burden_map_SEPTA <- tm_shape(SEPTA_2021) +
    tm_polygons(
      "inflation_adjusted_transit_cost_burden",
      palette = "viridis",
      title = "Total Transit Cost Burden 2021 - SEPTA",
      border.col = NULL
    ) +
    tm_layout(
      title.position = c("left", "top"),
      legend.position = c("right", "bottom"),
      frame = FALSE, # No external frame
      bg.color = "white",
      legend.text.size = 1, # 12-point font equivalent, adjust for visibility
      legend.title.size = 1.2,
      legend.outside = TRUE,
      title.color = "black"
    )

transit_burden_map_SEPTA <- transit_burden_map_SEPTA + tm_layout(
    fontfamily = "Bahnschrift",
    legend.title.fontface = "bold")


tmap_save(
  transport_burden_map_SEPTA,
  "transport_burden_map_SEPTA_li.png",
  width = 8,
  height = 6,
  dpi = 300
)
tmap_save(
  transit_burden_map_SEPTA,
  "transit_burden_map_SEPTA.png",
  width = 8,
  height = 6,
  dpi = 300
)
    
    
st(SEPTA_2021)

tmap_mode("plot")

tm_shape(data_2021) +
  tm_polygons(
    col = "inflation_adjusted_transit_cost_burden",
    pal = "viridis",
    title = "Transportation Cost Burdens 2021",
    border.col = "transparent"
    ) +
  tm_facets(by = "operator", ncol = 3) +
  tm_layout(legend.outside.size = 0.2)

```

# Social Vulnerability Analysis 


#Morans I Maps


```{r}
# Load packages
library(tmap)
library(dplyr)
library(sf)
library(openxlsx)
library(rgeoda)

# Ensure tmap is in plotting mode
tmap_mode("plot")



# Filter and split by operator + year
data_filtered <- full_data %>%
  filter(Year %in% c(2019, 2021)) %>%
  filter(!is.na(si_index), !is.na(transit_spending)) %>%
  group_by(operator, Year) %>%
  group_split()

# Compute clusters
processed_data <- lapply(data_filtered, function(df) {
  weights <- queen_weights(df, order = 1)
  bimoran <- local_bimoran(weights, df[c("log_si_index_mod", "log_transportation_cost_burden")])
  clusters <- lisa_clusters(bimoran)
  df$lisa_cluster <- factor(clusters, levels = 0:4,
                            labels = c("Not Significant", "High-High", "Low-Low", "Low-High", "High-Low"))
  return(df)
})

combined_data <- do.call(rbind, processed_data)

# Combine into one sf object
combined_data_clean <- combined_data %>%
  filter(!is.na(lisa_cluster)) %>%
  mutate(lisa_cluster = factor(
    lisa_cluster,
    levels = c("Not Significant", "High-High", "Low-Low", "Low-High", "High-Low")
  ))

# Matching color vector
cluster_colors <- c(
  "gray80",   # Not Significant
  "red",      # High-High
  "blue",     # Low-Low
  "lightblue",# Low-High
  "pink"      # High-Low
)


final_map <- 
  tm_basemap("CartoDB.PositronNoLabels") +
  tm_shape(combined_data_clean) +
  tm_fill(
    col = "lisa_cluster",
    palette = cluster_colors,
    border.col = NULL,
    legend.is.portrait = FALSE,
    title = "LISA Cluster - Log SI Index x Log Transportation Cost Burden"
    
  ) +
  tm_facets_grid(
    rows = "operator",
    columns = "Year",
    drop.units = TRUE,
    free.coords = TRUE
  ) +
  tm_layout(
    frame = FALSE,
    bg.color = "white",
    
    # Legend settings
    legend.outside = TRUE,
    legend.outside.position = "bottom",
    legend.align = "center",
    legend.text.color = "black",
    legend.text.size = 0.8,
    legend.title.size = 1,
    legend.title.fontface = "bold",
    legend.bg.color = NA,
    legend.bg.alpha = 0,

    # Panel labels
    panel.label.bg.color = NA,
    panel.label.frame = FALSE,
    panel.label.color = "black",
    panel.label.size = 1.5,
    panel.label.height = 1.2,
    panel.label.margin = unit(2, "lines"),

    # Layout
    asp = 0,
    outer.margins = 0,
    inner.margins = 0
  )

# Save high-resolution faceted map
tmap_save(
  final_map,
  filename = file.path("Faceted_Bivariate_Moran_Map_All_Operators_Transportation_Logged.pdf"),
  width = 10,
  height = 12,
  dpi = 1200
)



```

Morans I maps for each operator 

```{r}

library(tmap)
library(dplyr)
library(purrr)

# Make sure output folder exists
dir.create("individual_operator_maps", showWarnings = FALSE)

# Clean and prepare the data
combined_data_clean <- combined_data %>%
  filter(!is.na(lisa_cluster)) %>%
  mutate(lisa_cluster = factor(
    lisa_cluster,
    levels = c("Not Significant", "High-High", "Low-Low", "Low-High", "High-Low")
  ))

# Color palette
cluster_colors <- c(
  "gray80",   # Not Significant
  "red",      # High-High
  "blue",     # Low-Low
  "lightblue",# Low-High
  "pink"      # High-Low
)

# Get unique operators
operators <- unique(combined_data_clean$operator)

# Loop through each operator
walk(operators, function(op) {
  op_data <- combined_data_clean %>% filter(operator == op)

op_map <- 
  tm_basemap("CartoDB.PositronNoLabels") +
  tm_shape(op_data) +
  tm_fill(
    col = "lisa_cluster",
    palette = cluster_colors,
    border.col = NULL,
    legend.is.portrait = FALSE,
    title = paste(op, "LISA Cluster - Log SI Index x Log Transportation Cost Burden")
    ) +
  tm_facets(rows = "Year", drop.units = TRUE, free.coords = TRUE) +
  tm_layout(
    frame = FALSE,
    bg.color = "white",

    # Legend settings
    legend.outside = TRUE,
    legend.outside.position = "bottom",
    legend.text.color = "black",
    legend.text.size = 0.8,
    legend.title.size = 1,
    legend.title.fontface = "bold",
    legend.bg.color = NA,
    legend.bg.alpha = 0,

    # Panel labels
    panel.label.bg.color = NA,
    panel.label.frame = FALSE,
    panel.label.color = "black",
    panel.label.size = 1.5,
    panel.label.height = 1.2,
    panel.label.margin = unit(2, "lines"),

    # Layout
    asp = 0,
    outer.margins = 0,
    inner.margins = 0
  )

  filename <- paste0("Transport_Bivariate_Moran_Map_", gsub("[^A-Za-z0-9]", "_", op), ".pdf")

  tmap_save(
    op_map,
    filename = file.path("individual_operator_maps", filename),
    width = 10,
    height = 14,
    dpi = 600
  )
})


```


Summary Table: MBTA/SEPTA/Chicago(CTA/METRA/South Shore Line) 2019/2021

```{r}
library(dplyr)
library(gt)

summary_table_full_2019 <- full_data %>%
  filter(Year == 2019) %>%
  mutate(total_income = inflation_adjusted_mean_income * population) %>%
   st_drop_geometry() %>%
  group_by(operator, mode_group) %>%
  summarise(
    `Number of Tracts`  = n(), 
    `Total Population` = round(sum(population, na.rm = T), 0),
    `Average Income` = round(sum(total_income, na.rm = T)/sum(population, na.rm = T), 0),
    `Average Transit Miles` = round(mean(transit_miles, na.rm = TRUE), 0),
    `Average HH Transportation Costs` = round(mean(inflation_adjusted_transport_cost, na.rm = T),2),
    `Average HH Transit Cost` = round(mean(inflation_adjusted_transit_cost, na.rm = TRUE), 2),
    `Average Social Vulnerability Component Score_2019` = round(mean(si_index, na.rm = TRUE), 2)
  ) 

sv_table_19 <- summary_table_full_2019 %>%
  select(operator, mode_group, `Average Social Vulnerability Component Score_2019`) %>%
  pivot_wider(values_from = `Average Social Vulnerability Component Score_2019`, names_from = mode_group)



summary_table_full_2021 <- full_data %>%
    filter(Year == 2021) %>%
    mutate(total_income = inflation_adjusted_mean_income * population) %>%
   st_drop_geometry() %>%
 group_by(operator, mode_group) %>%
  summarise(
    `Number of Tracts` = n(),
    `Total Population` = round(sum(population, na.rm = T), 0),
    `Average Income` = round(sum(total_income, na.rm = T)/sum(population, na.rm=T), 0),
    `Average Transit Miles` = round(mean(transit_miles, na.rm = TRUE), 0),
    `Average HH Transportation Costs` = round(mean(inflation_adjusted_transport_cost, na.rm = T),0),
    `Average HH Transit Cost` = round(mean(inflation_adjusted_transit_cost, na.rm = TRUE), 0),
    `Average Social Vulnerability Component Score_2021` = round(mean(si_index, na.rm = TRUE), 2)
  )

sv_table_21 <- summary_table_full_2021 %>%
  select(operator, mode_group, `Average Social Vulnerability Component Score_2021`) %>%
  pivot_wider(values_from = `Average Social Vulnerability Component Score_2021`, names_from = mode_group)


difference_table <- summary_table_full_2019 %>%
  rename_with(~ paste0(.x, "_2019"), -c(mode_group, operator, `Average Social Vulnerability Component Score_2019`)) %>%
  inner_join(
    summary_table_full_2021 %>% 
      rename_with(~ paste0(.x, "_2021"), -c(mode_group, operator, `Average Social Vulnerability Component Score_2021`)),
    by = c("mode_group", "operator")
  ) %>%
  mutate(
    `Difference in Population` = `Total Population_2021` - `Total Population_2019`,
    `Difference in Average Income` = `Average Income_2021` - `Average Income_2019`,
    `Difference in Average Transit Miles` = `Average Transit Miles_2021` - `Average Transit Miles_2019`,
    `Difference in HH Transportation Costs` = `Average HH Transportation Costs_2021` - `Average HH Transportation Costs_2019`,
    `Difference in Average HH Transit Cost` = `Average HH Transit Cost_2021` - `Average HH Transit Cost_2019`,
    `Difference in Average Social Vulnerability Component Score` = `Average Social Vulnerability Component Score_2021` - `Average Social Vulnerability Component Score_2019`
  ) %>%
  select(
    mode_group, operator,
    "Number of Tracts" = `Number of Tracts_2019`,
    `Difference in Population`,
    `Difference in Average Income`,
    `Difference in Average Transit Miles`,
    `Difference in HH Transportation Costs`,
    `Difference in Average HH Transit Cost`,
    `Difference in Average Social Vulnerability Component Score`
  )




```
t tests
```{r}

t_test_mode_group_SEPTA <- data.frame()
t_test_mode_group_MBTA <- data.frame()
t_test_mode_group_LA <- data.frame()
t_test_mode_group_Chicago <- data.frame()
t_test_mode_group_MTA <- data.frame()
t_test_mode_group_MARTA <- data.frame()
t_test_mode_group_Seattle <- data.frame()

for(x in unique(full_data$mode_group)) {
  t_test_mode_group_SEPTA <-  rbind(
    t_test_mode_group_SEPTA,
    nice_t_test(
      full_data %>% filter(operator == "SEPTA", mode_group == x),
      response = c(
        "population",
        "inflation_adjusted_mean_income",
        "transit_miles",
        "inflation_adjusted_transport_cost",
        "inflation_adjusted_transit_cost",
        "si_index"
      ),
      group = "Year",
      warning = F
    ) %>%  mutate(mode_group = x)
  )
   
  
}

t_test_mode_group_SEPTA$operator <- "SEPTA"

for(x in unique(full_data$mode_group)) {
  t_test_mode_group_MBTA <-  rbind(
    t_test_mode_group_MBTA,
    nice_t_test(
      full_data %>% filter(operator == "MBTA", mode_group == x),
      response = c(
        "population",
        "inflation_adjusted_mean_income",
        "transit_miles",
        "inflation_adjusted_transport_cost",
        "inflation_adjusted_transit_cost",
        "si_index"
      ),
      group = "Year",
      warning = F
    ) %>%  mutate(mode_group = x)
  )
   
  
}

t_test_mode_group_MBTA$operator <- "MBTA"


for(x in unique(full_data$mode_group)) {
  t_test_mode_group_LA <-  rbind(
    t_test_mode_group_LA,
    nice_t_test(
      full_data %>% filter(operator == "LA", mode_group == x),
      response = c(
        "population",
        "inflation_adjusted_mean_income",
        "transit_miles",
        "inflation_adjusted_transport_cost",
        "inflation_adjusted_transit_cost",
        "si_index"
      ),
      group = "Year",
      warning = F
    ) %>%  mutate(mode_group = x)
  )
   
  
}

t_test_mode_group_LA$operator <- "LA"


for(x in unique(full_data$mode_group)) {
  t_test_mode_group_Chicago <-  rbind(
    t_test_mode_group_Chicago,
    nice_t_test(
      full_data %>% filter(operator == "Chicago", mode_group == x),
      response = c(
        "population",
        "inflation_adjusted_mean_income",
        "transit_miles",
        "inflation_adjusted_transport_cost",
        "inflation_adjusted_transit_cost",
        "si_index"
      ),
      group = "Year",
      warning = F
    ) %>%  mutate(mode_group = x)
  )
   
  
}

t_test_mode_group_Chicago$operator <- "Chicago"


for(x in unique(full_data$mode_group)) {
  t_test_mode_group_MTA <-  rbind(
    t_test_mode_group_MTA,
    nice_t_test(
      full_data %>% filter(operator == "MTA", mode_group == x),
      response = c(
        "population",
        "inflation_adjusted_mean_income",
        "transit_miles",
        "inflation_adjusted_transport_cost",
        "inflation_adjusted_transit_cost",
        "si_index"
      ),
      group = "Year",
      warning = F
    ) %>%  mutate(mode_group = x)
  )
   
  
}
t_test_mode_group_MTA$operator <- "MTA"



for(x in unique(full_data$mode_group)) {
  t_test_mode_group_Seattle <-  rbind(
    t_test_mode_group_Seattle,
    nice_t_test(
      full_data %>% filter(operator == "Seattle", mode_group == x),
      response = c(
        "population",
        "inflation_adjusted_mean_income",
        "transit_miles",
        "inflation_adjusted_transport_cost",
        "inflation_adjusted_transit_cost",
        "si_index"
      ),
      group = "Year",
      warning = F
    ) %>%  mutate(mode_group = x)
  )
   
  
}
t_test_mode_group_Seattle$operator <- "Seattle"



for(x in c("Bus and Rapid Transit Access", "Only Bus Access")) {
  t_test_mode_group_MARTA <-  rbind(
    t_test_mode_group_MARTA,
    nice_t_test(
      full_data %>% filter(operator == "MARTA", mode_group == x),
      response = c(
        "population",
        "inflation_adjusted_mean_income",
        "transit_miles",
        "inflation_adjusted_transport_cost",
        "inflation_adjusted_transit_cost",
        "si_index"
      ),
      group = "Year",
      warning = F
    ) %>%  mutate(mode_group = x)
  )
   
  
}
t_test_mode_group_MARTA$operator <- "MARTA"



t_tests_full <- rbind(t_test_mode_group_LA, t_test_mode_group_SEPTA, t_test_mode_group_Chicago, t_test_mode_group_MBTA,t_test_mode_group_MTA, t_test_mode_group_Seattle, t_test_mode_group_MARTA ) %>%
  mutate(Significance = ifelse(p > 0.05, "Not Significant", "Significant (at p<.05)"),
         tags = case_when(
           `Dependent Variable` == "population" ~ "Difference in Population",
           `Dependent Variable` == "inflation_adjusted_mean_income" ~ "Difference in Average Income",
           `Dependent Variable` == "transit_miles" ~ "Difference in Average Transit Miles",
           `Dependent Variable` == "inflation_adjusted_transport_cost" ~ "Difference in HH Transportation Costs",
           `Dependent Variable` == "inflation_adjusted_transit_cost" ~ "Difference in Average HH Transit Cost",
           `Dependent Variable` == "si_index" ~ "Difference in Average Social Vulnerability Component Score",
           TRUE ~ NA_character_  # Catch any unexpected values
         ))


# 
# difference_table_full <- difference_table %>%
#   pivot_longer(cols = -c(operator, mode_group), names_to = "tags", values_to = "value") %>%
#   left_join(t_tests_full %>% select(Significance, operator, mode_group, tags), 
#             by = c("operator", "mode_group", "tags"))
# 
# difference_table_wide <- difference_table_full %>%
#   pivot_wider(names_from = tags, values_from = value) %>%
#   pivot_wider(names_from = tags, values_from = Significance, names_prefix = "Significance_")



write.csv(t_tests_full,"t_tests_full.csv")


```
Nicely Formatted Tables
```{r}
library(flextable)


x <- difference_table %>%
  gt() %>%
  tab_header(
    title = "Difference Table: 2019 vs 2021",
    subtitle = "Changes in Key Metrics Between Years"
  ) %>%
  cols_label(
    mode_group = "Mode Group",
    operator = "Operator",
    `Number of Tracts` = "Number of Tracts",
    `Difference in Population` = "Diff: Total Population", 
    `Difference in Average Income` = "Diff: Average Income ($)",
    `Difference in Average Transit Miles` = "Diff: Average Transit Miles",
    `Difference in HH Transportation Costs` = "Diff: HH Transportation Costs ($)",
    `Difference in Average HH Transit Cost` = "Diff: HH Transit Cost ($)",
    `Difference in Average Social Vulnerability Component Score` = "Diff: Social Vulnerability Score"
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  fmt_number(
    columns = c(
      `Difference in Average Income`, 
      `Difference in HH Transportation Costs`, 
      `Difference in Average HH Transit Cost`
    ),
    use_seps = TRUE,  # Adds comma separators
    decimals = 0  # Removes decimal places for cleaner formatting
  ) %>%
  tab_options(
    table.font.size = 12,
    data_row.padding = px(6)
  ) %>%
   gtsave("difference.html")





```

Bar Plots Instead of Differnces
```{r}

plot_data <- full_data %>%
  st_drop_geometry() %>%
  group_by(mode_group, Year) %>%
  summarise(avg_transit_spending = mean(inflation_adjusted_transit_cost, na.rm = TRUE),
            avg_transport_spending = mean(inflation_adjusted_transport_cost, na.rm = T), 
            avg_si_index = mean(si_index, na.rm = TRUE))

plot_data_raw <- full_data %>%
  st_drop_geometry() %>%
  select(mode_group, Year, si_index) 

plot_data_raw_filtered <- plot_data_raw %>%
  filter(Year %in% c(2019, 2021))

# 1. Histogram for Transport Cost Burden (assuming you have a `transport_cost_burden` variable)
ggplot(full_data %>% st_drop_geometry(), aes(x = inflation_adjusted_transport_cost, fill = factor(Year))) +
  geom_histogram(binwidth = 0.1, position = "dodge", alpha = 0.7, color = "black") +
  geom_vline(aes(xintercept = mean(inflation_adjusted_transport_cost, na.rm = TRUE)), color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Transport Cost Burden (2019 & 2021)",
    x = "Transport Cost Burden",
    y = "Frequency"
  ) +
  scale_fill_brewer(palette = "Dark2") +  # Dark earth-tone palette
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, family = "Bahnschrift"),
    axis.text.y = element_text(family = "Bahnschrift"),
    legend.title = element_blank(),
    plot.title = element_text(family = "Bahnschrift")
  )

# 2. Histogram for Transit Cost Burden (assuming you have a `transit_cost_burden` variable)
ggplot(plot_data_raw_filtered, aes(x = transit_cost_burden, fill = factor(Year))) +
  geom_histogram(binwidth = 0.1, position = "dodge", alpha = 0.7, color = "black") +
  geom_vline(aes(xintercept = mean(transit_cost_burden, na.rm = TRUE)), color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Transit Cost Burden (2019 & 2021)",
    x = "Transit Cost Burden",
    y = "Frequency"
  ) +
  scale_fill_brewer(palette = "Dark2") +  # Dark earth-tone palette
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, family = "Bahnschrift"),
    axis.text.y = element_text(family = "Bahnschrift"),
    legend.title = element_blank(),
    plot.title = element_text(family = "Bahnschrift")
  )

# 3. Histogram for SI Index
ggplot(plot_data_raw_filtered, aes(x = si_index, fill = factor(Year))) +
  geom_histogram(binwidth = 0.1, position = "dodge", alpha = 0.7, color = "black") +
  geom_vline(aes(xintercept = mean(si_index, na.rm = TRUE)), color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "SI Index (2019 & 2021)",
    x = "SI Index",
    y = "Frequency"
  ) +
  scale_fill_brewer(palette = "Dark2") +  # Dark earth-tone palette
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, family = "Bahnschrift"),
    axis.text.y = element_text(family = "Bahnschrift"),
    legend.title = element_blank(),
    plot.title = element_text(family = "Bahnschrift")
  )


```

# Scatterplots
```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)

# Scatter plot for 2019
scatter_plot_2019_transit <- ggplot(
  full_data %>% filter(Year == 2019),
  aes(x = log_transit_cost_burden, y = si_index, color = mode_group)
) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = colors) +
  labs(
    title = "Log % Income Spent on Transit and Social Vulnerability (2019)",
    x = "Log % of HH Income Spent on Transit",
    y = "Social Vulnerability Index",
    color = "Mode Access"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Bahnschrift"),
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    panel.grid.minor = element_blank()
  ) 

# Scatter plot for 2021
scatter_plot_2021_transit <- ggplot(
  full_data %>% filter(Year == 2021),
  aes(x = log_transit_cost_burden, y = si_index, color = mode_group)
) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = colors) +
  labs(
    title = "Log % Income Spent on Transit and Social Vulnerability (2021)",
    x = "Log % of HH Income Spent on Transit",
    y = "Social Vulnerability Index",
    color = "Mode Access"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Bahnschrift"),
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    panel.grid.minor = element_blank()
  ) 

# Scatter plot for 2019
scatter_plot_2019_transport <- ggplot(
  full_data_mod %>% filter(Year == 2019),
  aes(x = log_transportation_cost_burden, y = si_index, color = mode_group)
) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = colors) +
  labs(
    title = "Log % Income Spent on Transportation and Social Vulnerability (2019)",
    x = "Log % of HH Income Spent on Transportation",
    y = "Social Vulnerability Index",
    color = "Mode Access"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Bahnschrift"),
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    panel.grid.minor = element_blank()
  ) 

# Scatter plot for 2021
scatter_plot_2021_transport <- ggplot(
  full_data_mod %>% filter(Year == 2021),
  aes(x = log_transportation_cost_burden, y = si_index, color = mode_group)
) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = colors) +
  labs(
    title = "Log % Income Spent on Transportation and Social Vulnerability (2021)",
    x = "Log % of HH Income Spent on Transportation",
    y = "Social Vulnerability Index",
    color = "Mode Access"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Bahnschrift"),
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    panel.grid.minor = element_blank()
  ) 

# Arrange scatter plots together
scatter_arrange_transport <- ggarrange(scatter_plot_2019_transport, scatter_plot_2021_transport, ncol = 2, common.legend = TRUE, legend = "bottom")


# Arrange scatter plots together
scatter_arrange_transit <- ggarrange(scatter_plot_2019_transit, scatter_plot_2021_transit, ncol = 2, common.legend = TRUE, legend = "bottom")



```

Line Plots
```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)
library(grid)  # for unit()
loadfonts(device = "pdf")
font_add("Bahnschrift", regular = "C:/Windows/Fonts/bahnschrift.ttf")

# Define custom theme with no legend title and adjusted spacing
custom_theme <- theme_minimal() +
  theme(
    text = element_text(family = "Bahnschrift"),
    axis.title = element_text(size = 15),
    axis.title.x = element_text(margin = margin(t = 12)),
    axis.text = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 15),
    legend.key.width = unit(1.5, "cm"),       # Wider key
    legend.box.spacing = unit(1.5, "cm"),     # Spacing between items
    legend.margin = margin(t = 15, r = 15, b = 15, l = 15),
    panel.grid.minor = element_blank()
  )

# Define guide for consistent legend layout
shared_legend_guide <- guide_legend(
  nrow = 2,
  byrow = TRUE,
  label.theme = element_text(margin = margin(l = 6, r = 6))
)

# Line plot for 2019 (Transit)
line_plot_2019_transit <- ggplot(
  full_data %>% filter(Year == 2019),
  aes(x = log_transit_cost_burden, y = si_index, color = mode_group)
) +
  geom_smooth(method = "lm", se = FALSE, size = 1.5) +
  scale_color_manual(values = colors, guide = shared_legend_guide) +
  labs(
    x = "Log % of HH Income Spent on Transit",
    y = "Social Vulnerability Index",
    color = NULL
  ) +
  custom_theme

# Line plot for 2021 (Transit)
line_plot_2021_transit <- ggplot(
  full_data %>% filter(Year == 2021),
  aes(x = log_transit_cost_burden, y = si_index, color = mode_group)
) +
  geom_smooth(method = "lm", se = FALSE, size = 1.5) +
  scale_color_manual(values = colors, guide = shared_legend_guide) +
  labs(
    x = "Log % of HH Income Spent on Transit",
    y = "Social Vulnerability Index",
    color = NULL
  ) +
  custom_theme

# Line plot for 2019 (Transportation)
line_plot_2019_transport <- ggplot(
  full_data_mod %>% filter(Year == 2019),
  aes(x = log_transportation_cost_burden, y = si_index, color = mode_group)
) +
  geom_smooth(method = "lm", se = FALSE, size = 1.5) +
  scale_color_manual(values = colors, guide = shared_legend_guide) +
  labs(
    x = "Log % of HH Income Spent on Transportation",
    y = "Social Vulnerability Index",
    color = NULL
  ) +
  custom_theme

# Line plot for 2021 (Transportation)
line_plot_2021_transport <- ggplot(
  full_data_mod %>% filter(Year == 2021),
  aes(x = log_transportation_cost_burden, y = si_index, color = mode_group)
) +
  geom_smooth(method = "lm", se = FALSE, size = 1.5) +
  scale_color_manual(values = colors, guide = shared_legend_guide) +
  labs(
    x = "Log % of HH Income Spent on Transportation",
    y = "Social Vulnerability Index",
    color = NULL
  ) +
  custom_theme

# Combine all 4 plots with shared legend
all_line_plots <- ggarrange(
  line_plot_2019_transport, line_plot_2021_transport,
  line_plot_2019_transit, line_plot_2021_transit,
  ncol = 2, nrow = 2,
  common.legend = TRUE,
  legend = "bottom"
)

# Save to PDF
ggsave(
  all_line_plots,
  filename = "combined_line_plots.pdf",
  width = 15, height = 18, dpi = 600
)
```

# models - Diff in Diff - All systems 2019 and 2021 - The Average Household

```{r Diff in Diff}

control_vars <- c("population", "log_emp_density")


mod1 <- lm(as.formula(
  paste(
    "log_si_index ~ log_transportation_cost_burden + cr_tracts + operator + Year + log_transportation_cost_burden*Year + log_transportation_cost_burden*Year*cr_tracts +",
    paste(control_vars, collapse = " + ")
  )),
  data = full_data 
)

mod2 <- lm(as.formula(
  paste(
    "log_si_index_mod  ~ log_transit_cost_burden + cr_tracts + operator + Year_2021 + log_transit_cost_burden*Year_2021 + log_transit_cost_burden*Year_2021*cr_tracts +", 
     paste(control_vars, collapse = " + ")
  )),
  data = full_data
)


#DID Analysis Models

library(sandwich)
library(lmtest)

# Fit your model as you normally would
mod3 <- lm(as.formula(
  paste(
    "log_transit_cost_burden ~ cr_tracts + Year_2021 + si_index_high + operator + cr_tracts*Year_2021 + si_index_high*cr_tracts*Year_2021 +",
    paste(control_vars, collapse = " + ")
  )),
  data = full_data_mod 
)

# Compute robust standard errors using vcovHC
robust_se <- vcovHC(mod3, type = "HC3")  # Type "HC3" is often recommended for small samples

# View the results with robust standard errors
coeftest(mod3, vcov. = robust_se)

mod4 <- lm(as.formula(
  paste(
    "log_transit_cost_burden ~ rt_tracts + Year_2021 + si_index_high + operator + rt_tracts*Year_2021 + si_index_high*rt_tracts*Year_2021 + ",
     paste(control_vars, collapse = " + ")
  )),
  data = full_data 
)

mod5 <- lm(as.formula(
  paste(
    "log_transit_cost_burden ~ bus_tracts + Year_2021 + si_index_high + operator + bus_tracts*Year_2021 + si_index_high*bus_tracts*Year_2021 + ",
     paste(control_vars, collapse = " + ")
  )),
  data = full_data 
)

mod6 <- lm(as.formula(
  paste(
    "log_transit_cost_burden ~ cr_tracts_only + Year_2021 + si_index_high + operator + cr_tracts_only*Year_2021 + si_index_high*cr_tracts_only*Year_2021 + ",
     paste(control_vars, collapse = " + ")
  )),
  data = full_data 
)


# Load the car package for vif()
library(car)

# Create the formula string including your controls
formula_str <- paste(
  "log_transportation_cost_burden ~ treatment * Year_2021 * si_index_high + operator +",
  paste(control_vars, collapse = " + ")
)

# Fit the linear model
lm_transport <- lm(as.formula(formula_str), data = full_data_mod)

# Calculate VIFs
vif_values <- vif(lm_transport)

# Print VIF values
print(vif_values)




```


#Spatial Regression

```{r}
library(spdep)
library(spatialreg)

# Compute Moran's I
neighbors <- poly2nb(full_data_mod, queen = TRUE)
weights   <- nb2listw(neighbors, style = "W", zero.policy = TRUE)


#morans tests indicates high spatial autocorrelation 
moran_test <- moran.test(full_data_mod$log_transportation_cost_burden, weights, zero.policy = TRUE)

#lm tests indicates that i need to use an error model 
# lm_tests <- lm.LMtests(mod3, weights, test = "all")


# Spatial Error and Lag Models 



#CR Tracts as Treatment

full_data_mod$treatment <- as.factor(full_data_mod$cr_tracts)

# transit
error_model_cr <- errorsarlm(as.formula(
  paste(
    "log_transit_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")


lag_model_cr <- lagsarlm(
  as.formula(
    paste(
      "log_transit_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
      paste(control_vars, collapse = " + ")
    )
  ),
  data = full_data_mod,
  listw = weights,
  method = "Matrix"
)

#transport

error_model_cr_transport <- errorsarlm(as.formula(
  paste(
    "log_transportation_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")


lag_model_cr_transport <- lagsarlm(
  as.formula(
    paste(
      "log_transportation_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
      paste(control_vars, collapse = " + ")
    )
  ),
  data = full_data_mod,  # Exclude Seattle for this model
  listw = weights,
  method = "Matrix"
)


# Tests to choose between spatial error or lag - should choose lag (lower AIC/BIC and less spatial autocorrelation)
AIC(error_model_cr, lag_model_cr)
BIC(error_model_cr, lag_model_cr)

moran.test(residuals(error_model_cr), weights)
moran.test(residuals(lag_model_cr), weights)


residuals_cr <- residuals(error_model_cr)
fitted_values_cr <- fitted(error_model_cr)

# Base R plot (simple)
plot(fitted_values_cr, residuals_cr, 
     main = "Residual Plot", 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     pch = 16, col = "blue")
abline(h = 0, col = "red")  # Add a horizontal line at 0


#RT Tracts

full_data_mod$treatment <- as.factor(full_data_mod$rt_tracts)

# transit
error_model_rt <- errorsarlm(as.formula(
  paste(
    "log_transit_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")


lag_model_rt <- lagsarlm(
  as.formula(
    paste(
      "log_transit_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
      paste(control_vars, collapse = " + ")
    )
  ),
  data = full_data_mod,
  listw = weights,
  method = "Matrix"
)

# transport
error_model_rt_transport <- errorsarlm(as.formula(
  paste(
    "log_transportation_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")


lag_model_rt_transport <- lagsarlm(
  as.formula(
    paste(
      "log_transportation_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
      paste(control_vars, collapse = " + ")
    )
  ),
  data = full_data_mod,
  listw = weights,
  method = "Matrix"
)

#Bus Tracts

full_data_mod$treatment <- as.factor(full_data_mod$bus_tracts)

#transit
error_model_bus <- errorsarlm(as.formula(
  paste(
    "log_transit_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")

lag_model_bus <- lagsarlm(
  as.formula(
    paste(
      "log_transit_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
      paste(control_vars, collapse = " + ")
    )
  ),
  data = full_data_mod,
  listw = weights,
  method = "Matrix"
)

#transport

error_model_bus_transport <- errorsarlm(as.formula(
  paste(
    "log_transportation_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")

lag_model_bus_transport <- lagsarlm(
  as.formula(
    paste(
      "log_transportation_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
      paste(control_vars, collapse = " + ")
    )
  ),
  data = full_data_mod,
  listw = weights,
  method = "Matrix"
)

#CR Tracts Only

full_data_mod$treatment <- as.factor(full_data_mod$cr_tracts_only)

#transit
error_model_cr_only <- errorsarlm(as.formula(
  paste(
    "log_transit_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")

lag_model_cr_only <- lagsarlm(
  as.formula(
    paste(
      "log_transit_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
      paste(control_vars, collapse = " + ")
    )
  ),
  data = full_data_mod,
  listw = weights,
  method = "Matrix"
)

#transport
error_model_cr_only_transport <- errorsarlm(as.formula(
  paste(
    "log_transportation_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")

lag_model_cr_only_transport <- lagsarlm(
  as.formula(
    paste(
      "log_transportation_cost_burden ~ treatment + Year_2021 + si_index_high + operator + treatment*Year_2021 + si_index_high*treatment*Year_2021 + ",
      paste(control_vars, collapse = " + ")
    )
  ),
  data = full_data_mod,
  listw = weights,
  method = "Matrix"
)

table(full_data_mod$treatment, full_data_mod$Year_2021, full_data_mod$si_index_high)


#Switched - CR Tracts Only Predicting Log Social Vulnerability 

full_data_mod$treatment <- full_data_mod$log_transit_cost_burden

error_model_cr_only_si_index_transit <- errorsarlm(as.formula(
  paste(
    "log_si_index  ~ treatment + cr_tracts_only + operator + Year_2021 + cr_tracts_only*Year_2021 + treatment*Year_2021*cr_tracts_only +",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")


error_model_cr_si_index_transit <- errorsarlm(as.formula(
  paste(
    "log_si_index  ~ treatment + cr_tracts + operator + Year_2021 + cr_tracts*Year_2021 + treatment*Year_2021*cr_tracts +",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")


moran.test(residuals(error_model_cr_only_si_index_transit), weights)
# Going with error model here because of lower AIC/BIC


full_data_mod$treatment <- full_data_mod$log_transportation_cost_burden


error_model_cr_only_si_index_transport <- errorsarlm(as.formula(
  paste(
    "log_si_index  ~ treatment + cr_tracts_only + operator + Year_2021 + cr_tracts_only*Year_2021 + treatment*Year_2021*cr_tracts_only +",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")



error_model_cr_si_index_transport <- errorsarlm(as.formula(
  paste(
    "log_si_index  ~ treatment + cr_tracts + operator + Year_2021 + cr_tracts*Year_2021 + treatment*Year_2021*cr_tracts +",
    paste(control_vars, collapse = " + ")
  )
), data = full_data_mod, listw = weights, method = "Matrix")


 

```




```{r}


library(stargazer)


#transit cost burden

# Extract lag coefficients and standard errors
lag_coefs <- sapply(
  list(lag_model_cr, lag_model_rt, lag_model_bus, lag_model_cr_only),
  function(model) model$rho
)

lag_se <- sapply(
  list(lag_model_cr, lag_model_rt, lag_model_bus, lag_model_cr_only),
  function(model) sqrt(model$rho.se)
)

# Compute z-scores and p-values
lag_z <- lag_coefs / lag_se
lag_p <- 2 * (1 - pnorm(abs(lag_z)))

# Significance stars
sig_stars <- ifelse(lag_p < 0.001, "***",
              ifelse(lag_p < 0.01, "**",
              ifelse(lag_p < 0.05, "*",
              ifelse(lag_p < 0.1, ".", ""))))

# Format rho with SE and significance
rho_row <- paste0(
  round(lag_coefs, 3), " (", round(lag_se, 3), ")", sig_stars
)

# Add this to your stargazer table
html_output <- stargazer(
  lag_model_cr,
  lag_model_rt,
  lag_model_bus,
  lag_model_cr_only,
  type = "latex",
  title = "Spatial Lag Models for Transit Cost Burden",
  dep.var.labels.include = TRUE,
  dep.var.labels = "Log Transit Cost Burden",
  column.labels = c(
    "Commuter Rail Access",
    "Rapid Transit Access",
    "Bus Access",
    "Commuter Rail Access ONLY"
  ),
  covariate.labels = c(
    "Mode Access Treatment",
    "Year 2021",
    "SI Index High", 
    "Treatment * Year 2021",
    "Treatment * SI Index High",
    "Year 2021 * SI Index High",
    "SI Index High * Treatment * Year 2021"
  ),
  add.lines = list(c("Lag Coefficient (rho)", rho_row)),
  omit.stat = c("f", "ser"),
  omit = c("operator", "population", "log_emp_density"),
  notes = "Standard errors are robust to spatial correlation. Operator, population, and employment density are included as controls but omitted from the table.",
  notes.align = "l"
)

# Save as an HTML file
#write(html_output, file = "spatial_lag_models_transit_cost_burden.html")
write(html_output, file = "spatial_lag_models_transit_cost_burden.tex")

#SI index 
html_output_2 <- stargazer(
  error_model_cr_only_si_index_transit,
  error_model_cr_only_si_index_transport,
  type = "latex",
  # Output as HTML
  title = "Spatial Error Models: Cost Burden Effect on SI Index (Commuter Rail Only)",
  dep.var.labels = "Log SI Index",
  column.labels = c(
    "Transit Cost Burden",
    "Transport Cost Burden"
  ),
  covariate.labels = c(
    "Log Cost Burden", 
    "Mode Access Treatment - CR Tracts Only",
    "Year 2021",
    "CR Tracts Only * Year 2021",
    "Log Cost Burden * Year 2021",
    "CR Tracts Only * Log Cost Burden",
    "Log Cost Burden * CR Tracts Only * Year 2021"
  ),
  omit.stat = c("f", "ser"),
  omit = c("operator", "population", "log_emp_density"),
  notes = "Standard errors are robust to spatial correlation.Operator, population, and employment density are included as controls but omitted from the table.",
  notes.align = "l"
)

#write(html_output_2, file = "spatial_error_models_si_index.html")
write(html_output_2, file = "spatial_error_models_si_index.tex")

# transport cost burden

# Extract lag coefficients and standard errors
# lag_coefs_transport <- sapply(
#   list(lag_model_cr_transport, lag_model_rt_transport, lag_model_bus_transport, lag_model_cr_only_transport),
#   function(model) model$rho
# )
# 
# lag_se_transport <- sapply(
#   list(lag_model_cr_transport, lag_model_rt_transport, lag_model_bus_transport, lag_model_cr_only_transport),
#   function(model) sqrt(model$rho.se)
# )
# 
# # Compute z-scores and p-values
# lag_z_transport <- lag_coefs_transport / lag_se_transport
# lag_p_transport <- 2 * (1 - pnorm(abs(lag_z_transport)))
# 
# # Significance stars
# sig_stars_transport <- ifelse(lag_p_transport < 0.001, "***",
#               ifelse(lag_p_transport < 0.01, "**",
#               ifelse(lag_p_transport < 0.05, "*",
#               ifelse(lag_p_transport < 0.1, ".", ""))))
# 
# # Format rho with SE and significance
# rho_row_transport <- paste0(
#   round(lag_coefs_transport, 3), " (", round(lag_se_transport, 3), ")", sig_stars_transport
# )
# 
# # Add this to your stargazer table
# html_output_transport <- stargazer(
#   lag_model_cr_transport,
#   lag_model_rt_transport,
#   lag_model_bus_transport,
#   lag_model_cr_only_transport,
#   type = "html",
#   title = "Spatial Lag Models for Transport Cost Burden",
#   dep.var.labels.include = TRUE,
#   dep.var.labels = "Log Transport Cost Burden",
#   column.labels = c(
#     "Commuter Rail Access",
#     "Rapid Transit Access",
#     "Bus Access",
#     "Commuter Rail Access ONLY"
#   ),
#   covariate.labels = c(
#     "Mode Access Treatment",
#     "Year 2021",
#     "SI Index High", 
#     "Treatment * Year 2021",
#     "Treatment * SI Index High",
#     "Year 2021 * SI Index High",
#     "SI Index High * Treatment * Year 2021"
#   ),
#   add.lines = list(c("Lag Coefficient (rho)", rho_row_transport)),
#   omit.stat = c("f", "ser"),
#   omit = c("operator", "population", "log_emp_density"),
#   notes = "Standard errors are robust to spatial correlation. Operator, population, and employment density are included as controls but omitted from the table.",
#   notes.align = "l"
# )
# 
# write(html_output_transport, file = "spatial_lag_models_transport_cost_burden.html")
# 


```





